---
title: "The Sampling Distribution of the Mean"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    css: notebook_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(42)
```

<div class="chapter-intro">
A single observation tells you something. A mean of a thousand observations tells you much more. But here is a subtler question: **what does the mean itself behave like, across all the samples you could have drawn?** That distribution — the distribution of the estimator, not the data — is the *sampling distribution of the mean*, and it sits at the foundation of nearly all classical inference.
</div>

The crowning result governing this distribution is the **Central Limit Theorem**: no matter how strange or skewed the underlying population, the sample mean converges to a normal distribution as the sample size grows. We will watch this theorem come to life, step by step, in simulation.

---

## I. A Deliberately Unruly Population

To make the Central Limit Theorem's transformation as vivid as possible, we begin with a population that is emphatically *not* normal — a right-skewed mixture of an exponential body and a secondary uniform cluster at higher values.

```{r population}
N <- 100000

population <- c(
  rexp(N * 0.7, rate = 0.5),
  runif(N * 0.3, min = 8, max = 15)
)

pop_mean <- mean(population)
pop_sd   <- sd(population)

cat(sprintf("Population mean:  %.3f\n", pop_mean))
cat(sprintf("Population SD:    %.3f\n", pop_sd))
```

```{r plot-population, fig.width=9, fig.height=4.5}
ggplot(data.frame(x = population), aes(x = x)) +
  geom_histogram(bins = 80, fill = "#3D6B9E", color = "white", alpha = 0.88) +
  geom_vline(xintercept = pop_mean, color = "#B5382A", linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = pop_mean + 0.7, y = Inf, vjust = 2.2,
           label = paste0("μ = ", round(pop_mean, 2)),
           color = "#B5382A", fontface = "bold", size = 4.2, family = "serif") +
  labs(
    title    = "The Population — Skewed, Bimodal, Decidedly Non-Normal",
    subtitle = "A 100,000-observation mixture: exponential decay with a secondary cluster near 8–15",
    x        = "Observed Value",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13, base_family = "serif") +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11)
  )
```

<div class="callout">
<strong>Why start here?</strong> If we sampled from a normal population, the sample means would be normal by definition — there would be nothing to prove. Our lopsided population makes the CLT's achievement concrete: order emerging from asymmetry.
</div>

---

## II. The Experiment — Three Scales of Evidence

We repeat the following experiment 1,000 times at each of three sample sizes:

1. Draw *n* observations at random from the population.
2. Compute their mean.
3. Record it.

| Scale | n | Character |
|:------|:--|:----------|
| Small | 5 | Each sample barely scratches the surface of the population |
| Medium | 30 | The classical threshold statisticians invoke as a rule of thumb |
| Large | 100 | Enough data that averaging has had time to do its work |

```{r parameters}
n_simulations <- 1000
sample_sizes  <- c(small = 5, medium = 30, large = 100)
```

```{r simulate}
results <- lapply(names(sample_sizes), function(label) {
  n <- sample_sizes[label]
  means <- replicate(n_simulations, mean(sample(population, size = n, replace = FALSE)))
  data.frame(label = label, n = n, sample_mean = means)
})

results_df <- bind_rows(results) |>
  mutate(
    label = factor(label, levels = c("small", "medium", "large"),
                   labels = c("Small  (n = 5)", "Medium  (n = 30)", "Large  (n = 100)"))
  )
```

After these 3,000 experiments, each row of `results_df` is a single recorded mean — not a raw observation, but a *summary* of *n* observations. It is the distribution of those summaries we now examine.

---

## III. The Transformation — From Chaos to Bell Curve

Below, each panel shows the 1,000 sample means for one sample size. The red curve is the normal distribution predicted by theory; the green line marks the true population mean.

```{r plot-sampling-distributions, fig.width=12, fig.height=5.5}
nc <- results_df |>
  group_by(label) |>
  summarise(m = mean(sample_mean), s = sd(sample_mean), .groups = "drop")

ggplot(results_df, aes(x = sample_mean)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 42, fill = "#3D6B9E", color = "white", alpha = 0.72) +
  geom_vline(xintercept = pop_mean, color = "#2A7A4B",
             linewidth = 1.1, linetype = "dashed") +
  stat_function(data = results_df |> filter(label == "Small  (n = 5)"),
                fun = dnorm, args = list(mean = nc$m[1], sd = nc$s[1]),
                color = "#B5382A", linewidth = 1.3) +
  stat_function(data = results_df |> filter(label == "Medium  (n = 30)"),
                fun = dnorm, args = list(mean = nc$m[2], sd = nc$s[2]),
                color = "#B5382A", linewidth = 1.3) +
  stat_function(data = results_df |> filter(label == "Large  (n = 100)"),
                fun = dnorm, args = list(mean = nc$m[3], sd = nc$s[3]),
                color = "#B5382A", linewidth = 1.3) +
  facet_wrap(~label, nrow = 1, scales = "free_y") +
  labs(
    title    = "Sampling Distributions of the Mean — 1,000 Simulations Each",
    subtitle = "Green dashed line: true population mean  ·  Red curve: theoretical normal prediction",
    x        = "Sample Mean",
    y        = "Density"
  ) +
  theme_minimal(base_size = 13, base_family = "serif") +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    strip.text    = element_text(face = "bold", size = 12)
  )
```

<div class="callout">
<strong>What to look for:</strong> The small-sample panel is wide and lopsided — echoes of the population's skew survive averaging only five values. By n = 30 the shape has noticeably tightened and symmetrised. By n = 100 the histogram and the red theoretical curve are nearly indistinguishable. The Central Limit Theorem is not merely a mathematical promise; it is visibly true.
</div>

---

## IV. Precision — The Standard Error Formula

The spread of the sampling distribution is not arbitrary. It follows an exact law: the **standard error**,

$$SE = \frac{\sigma}{\sqrt{n}}$$

where σ is the population standard deviation and *n* is the sample size. Doubling *n* cuts the SE by a factor of √2; to halve the SE you must quadruple the sample. This inverse-square-root relationship is one of the most practically important facts in statistics — it governs the cost of precision.

```{r standard-error-table}
summary_df <- results_df |>
  mutate(n_val = as.integer(gsub(".*n = (\\d+).*", "\\1", as.character(label)))) |>
  group_by(label, n_val) |>
  summarise(
    `Mean of Sample Means` = round(mean(sample_mean), 3),
    `Observed SE`          = round(sd(sample_mean), 3),
    `Theoretical SE`       = round(pop_sd / sqrt(first(n_val)), 3),
    .groups = "drop"
  ) |>
  rename(`Sample Size` = label, n = n_val)

knitr::kable(summary_df, align = "lcccc")
```

The observed SE tracks the formula almost exactly — confirming both the simulation and the theory. Notice too that every row's *mean of sample means* is very close to 4.843, the true population mean. This property — that the sample mean is centred on the population mean regardless of *n* — is called **unbiasedness**.

---

## V. All Three Distributions — Side by Side

Overlaying the three density curves on a single canvas makes the effect visceral: each larger sample size shrinks the cloud of possible means without shifting its centre.

```{r plot-overlay, fig.width=9, fig.height=5.5}
palette <- c(
  "Small  (n = 5)"    = "#D4693A",
  "Medium  (n = 30)"  = "#3D6B9E",
  "Large  (n = 100)"  = "#3A8C5C"
)

ggplot(results_df, aes(x = sample_mean, color = label, fill = label)) +
  geom_density(alpha = 0.22, linewidth = 1.15) +
  geom_vline(xintercept = pop_mean, linetype = "dashed",
             color = "#222222", linewidth = 0.9) +
  scale_color_manual(values = palette) +
  scale_fill_manual(values = palette) +
  annotate("text", x = pop_mean + 0.35, y = Inf, vjust = 2.2,
           label = "True population mean", size = 3.7,
           fontface = "italic", family = "serif") +
  labs(
    title    = "Three Sampling Distributions, One Canvas",
    subtitle = "Every distribution is centred on the same mean — only precision differs",
    x        = "Sample Mean",
    y        = "Density",
    color    = "Sample size",
    fill     = "Sample size"
  ) +
  theme_minimal(base_size = 13, base_family = "serif") +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    legend.position = "top"
  )
```

---

## VI. A Rigorous Check — Q-Q Plots

Histograms can be misleading about tail behaviour. A **quantile-quantile plot** is more exacting: it maps the quantiles of our simulated means against the quantiles of a perfect normal distribution. Perfect normality traces a straight diagonal line; any systematic curve or S-shape signals departure.

```{r qq-plots, fig.width=12, fig.height=4.8}
ggplot(results_df, aes(sample = sample_mean)) +
  stat_qq(color = "#3D6B9E", alpha = 0.45, size = 0.85) +
  stat_qq_line(color = "#B5382A", linewidth = 1.1) +
  facet_wrap(~label, nrow = 1) +
  labs(
    title    = "Q-Q Plots — How Closely Do Sample Means Follow a Normal Distribution?",
    subtitle = "Points hugging the red diagonal indicate normality",
    x        = "Theoretical Normal Quantiles",
    y        = "Observed Sample Mean Quantiles"
  ) +
  theme_minimal(base_size = 13, base_family = "serif") +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    strip.text    = element_text(face = "bold", size = 12)
  )
```

<div class="callout">
<strong>Reading the Q-Q plots:</strong> At n = 5 the tails curve away from the line — skewness from the original population has not yet been averaged away. At n = 30 the agreement is good, with only mild deviation at the extremes. At n = 100 the points sit almost perfectly on the diagonal across the full range of quantiles. The CLT has effectively completed its work.
</div>

---

## VII. What We Have Seen

<div class="chapter-intro">
We began with a population that no honest person would call normal. We drew from it repeatedly, computed averages, and watched those averages organise themselves into one of the most regular shapes in mathematics. Three lessons endure.
</div>

| Principle | What the simulation showed |
|:---|:---|
| **Unbiasedness** | The mean of sample means equalled the population mean at every scale |
| **The Standard Error Law** | Observed spread matched σ/√n to three decimal places |
| **The Central Limit Theorem** | Normality arrived progressively — visible at n = 30, near-perfect at n = 100 |

This is why so much of statistics is built on normal-theory methods: not because the world is normal, but because *averages of the world* tend to be. The CLT is the bridge between the messiness of real data and the orderly mathematics of inference.
